---
title: "CCVI across U.S. Regions"
author: "Simon Minasse"
date: "Published 9/28/2020"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
library(tidyverse)
library(readxl)
library(RColorBrewer)
library(wesanderson)
library(ggpubr)

library(shiny)
```

```{r, include = FALSE}
#### Cleaning Data ####

# importing CCVI Surgo Dataset
state_data <- read_excel(path = "COVID-19 Community Vulnerability Index (CCVI)_V1.xlsx", sheet = 2)

glimpse(state_data)

# renaming columns to make referenceing easier and saving into new dataset
state_data2 <- state_data %>% rename(state = "State", state_abb = 'State Abbreviation', 
                                     theme_1 = "THEME 1: Socioeconomic Status",
                                     theme_2 = "THEME 2:\nHousehold Composition & Disability",
                                     theme_3 = "THEME 3: \nMinority Status & Language",
                                     theme_4 = "THEME 4: \nHousing Type & Transportation",
                                     theme_5 = "THEME 5: Epidemiological Factors",
                                     theme_6 = "THEME 6: Healthcare System Factors",
                                     ccvi_score = "CCVI SCORE\nHigher = More Vulnerable")
glimpse(state_data2) 


#### Defining Regions
# Redundant: Including Abbreviations in case it will be useful later

## West
west_rg <- c("ALASKA", "HAWAII", "WASHINGTON", "OREGON", "IDAHO", "MONTANA", "WYOMING", "CALIFORNIA", 
             "NEVADA", "UTAH", "COLORADO")
west_rg <- west_rg %>% as.data.frame() %>% rename(state = ".") %>% add_column(region = "WEST")

west_abb <- c("AK", "HI", "WA", "OR", "ID", "MT", "WY", "CA", "NV", "UT", "CO")
west_abb <- west_abb %>% as.data.frame() %>% rename(state_abb = ".") %>% add_column(region = "WEST")

## Southeast
southeast_rg <- c("ARKANSAS", "LOUISIANA", "MISSISSIPPI", "ALABAMA", "GEORGIA",
                  "FLORIDA", "SOUTH CAROLINA", "NORTH CAROLINA", "TENNESSEE",
                  "KENTUCKY", "WEST VIRGINIA", "VIRGINIA", "DISTRICT OF COLUMBIA", "MARYLAND",
                  "DELAWARE")
southeast_rg <- southeast_rg %>% as.data.frame() %>% rename(state = ".") %>% add_column(region = "SOUTHEAST")
southeast_abb <- c("AR", "LA", "MS", "AL", "GA", "FL", "SC", "NC",
                   "TN", "KY", "WV", "VA", "DC", "MD", "DE")
southeast_abb <- southeast_abb %>% as.data.frame() %>% rename(state_abb = ".") %>% add_column(region = "SOUTHEAST")

## Southwest
southwest_rg <- c("ARIZONA", "NEW MEXICO", "TEXAS", "OKLAHOMA")
southwest_rg <- southwest_rg %>% as.data.frame() %>% rename(state = ".") %>% add_column(region = "SOUTHWEST")

southwest_abb <- c("AZ", "NM", "TX", "OK")
southwest_abb <- southwest_abb %>% as.data.frame() %>% rename(state_abb = ".") %>% add_column(region = "SOUTHWEST")

## Midwest
midwest_rg <- c("NORTH DAKOTA", "SOUTH DAKOTA", "NEBRASKA", "KANSAS", 
                "MINNESOTA", "IOWA", "MISSOURI", "WISCONSIN", "ILLINOIS", 
                "INDIANA", "MICHIGAN", "OHIO")
midwest_rg <- midwest_rg %>% as.data.frame() %>% rename(state = ".") %>% add_column(region = "MIDWEST")

midwest_abb <- c("ND", "SD", "NE", "KS", "MD", "IA", "MO", "WI", "IL", "IN", "MI", "OH")
midwest_abb <- midwest_abb %>% as.data.frame() %>% rename(state_abb = ".") %>% add_column(region = "MIDWEST")


## Northeast
northeast_rg <- c("PENNSYLVANIA", "NEW JERSEY", "NEW YORK", "CONNECTICUT",
                  "RHODE ISLAND", "MASSACHUSETTS", "VERMONT", "NEW HAMPSHIRE",
                  "MAINE")
northeast_rg <- northeast_rg %>% as.data.frame() %>% rename(state = ".") %>% add_column(region = "NORTHEAST")

northeast_abb <- c("PA", "NJ", "NY", "CT", "RI", "MA", "VT", "NH", "ME")
northeast_abb <- northeast_abb %>% as.data.frame() %>% rename(state_abb = ".") %>% add_column(region = "NORTHEAST")


#### Stacking/concatenating and merging regional data with state data

# stacking regional data
state_info <- as.data.frame(rbind(midwest_rg, 
southwest_rg,
southeast_rg,
northeast_rg,
west_rg))

state_abb_info <- as.data.frame(rbind(midwest_abb, 
                                  southwest_abb,
                                  southeast_abb,
                                  northeast_abb,
                                  west_abb))

glimpse(state_info)
glimpse(state_abb_info)
glimpse(state_data2)

# creating new dataset that merges state and regional data 
state_data3 <- left_join(x = state_data2, y = state_info, by = 'state')
glimpse(state_data3)
state_data3
```

```{r, include = FALSE}
#### preparing data for plotting ####
state_data4 <- state_data3 %>% as.data.frame()
glimpse(state_data4)
head(state_data4)

# rearranging data for side-by-side (mean / median) plots
rg_mean <- state_data4 %>% group_by(region) %>% 
            summarise_if(is.numeric, mean) %>% 
            add_column(value = 'mean') %>% mutate_if(is.numeric, round, 2) 

rg_median <- state_data4 %>% group_by(region) %>% 
              summarise_if(is.numeric, median) %>%
              add_column(value = 'median') 

rg_mean; rg_median

rg_data <- rbind(rg_mean, rg_median)

rg_data <- rg_data %>% mutate_if(is.character, as.factor); rg_data

# plot ccvi mean / median side-by-side
p0 <- ggplot(data = rg_data, aes(x = region, y = ccvi_score, fill = value)) + 
  geom_col(position = position_dodge(), color = 'black') + 
  scale_fill_brewer(palette="Blues") +
  geom_text(aes(label=ccvi_score), position = position_dodge(width = 0.9),
            vjust=-0.5, size=4) +
  theme_minimal() +
  ylim(0,1) +
  ggtitle('Mean / Median CCVI Scores across U.S Regions')

# plot Theme 1 mean / median side-by-side
p1 <- ggplot(data = rg_data, aes(x = region, y = theme_1, fill = value)) + 
  geom_col(position = position_dodge(), color = 'black') + 
  scale_fill_brewer(palette="Greens") +
  geom_text(aes(label=theme_1), position = position_dodge(width = 0.9),
            vjust=-0.5, size=4) +
  theme_minimal() +
  ylim(0,1) +
  ggtitle('Mean / Median Socioeconomic Scores across U.S Regions')

# plot Theme 2 mean / median side-by-side
p2 <- ggplot(data = rg_data, aes(x = region, y = theme_2, fill = value)) + 
  geom_col(position = position_dodge(), color = 'black') + 
  scale_fill_brewer(palette="Greens") +
  geom_text(aes(label=theme_2), position = position_dodge(width = 0.9),
            vjust=-0.5, size=4) +
  theme_minimal() +
  ylim(0,1) +
  ggtitle('Mean / Median Household Scores across U.S Regions')

# plot Theme 3 mean / median side-by-side
p3 <- ggplot(data = rg_data, aes(x = region, y = theme_3, fill = value)) + 
  geom_col(position = position_dodge(), color = 'black') + 
  scale_fill_brewer(palette="Greens") +
  geom_text(aes(label=theme_3), position = position_dodge(width = 0.9),
            vjust=-0.5, size=4) +
  theme_minimal() +
  ylim(0,1) +
  ggtitle('Mean / Median Minority Scores across U.S Regions')

# plot Theme 4 mean / median side-by-side
p4 <- ggplot(data = rg_data, aes(x = region, y = theme_4, fill = value)) + 
  geom_col(position = position_dodge(), color = 'black') + 
  scale_fill_brewer(palette="Greens") +
  geom_text(aes(label=theme_4), position = position_dodge(width = 0.9),
            vjust=-0.5, size=4) +
  theme_minimal() +
  ylim(0,1) +
  ggtitle('Mean / Median Housing/Transportation Scores across U.S Regions')

# plot Theme 5 mean / median side-by-side
p5 <- ggplot(data = rg_data, aes(x = region, y = theme_5, fill = value)) + 
  geom_col(position = position_dodge(), color = 'black') + 
  scale_fill_brewer(palette="Greens") +
  geom_text(aes(label=theme_5), position = position_dodge(width = 0.9),
            vjust=-0.5, size=4) +
  theme_minimal() +
  ylim(0,1) +
  ggtitle('Mean / Median Epidemiological Scores across U.S Regions')

# plot Theme 6 mean / median side-by-side
p6 <- ggplot(data = rg_data, aes(x = region, y = theme_6, fill = value)) + 
  geom_col(position = position_dodge(), color = 'black') + 
  scale_fill_brewer(palette="Greens") +
  geom_text(aes(label=theme_6), position = position_dodge(width = 0.9),
            vjust=-0.5, size=4) +
  theme_minimal() +
  ylim(0,1) +
  ggtitle('Mean / Median Healthcare Scores across U.S Regions')

#### arranging plots on one page
ggarrange(p1,p2,p3,p4,p5,p6,p0, ncol = 2, nrow = 4)
```

##### Introduction
* Over the past few months I have come across a lot of interesting information and 
resources relating to the current pandemic. One of the things I have noticed in particular is the amount
of publicly accessible data relating to COVID-19.

* An interesting dataset I chose to work on was the [Surgo Foundation's COVID-19 Community Vulnerability Index (CCVI)][1]. The CCVI is a resource that helps to identify areas in the U.S. where support is most needed. As they put it, 
  "The CCVI is not designed to predict which individuals will become infected with coronavirus – instead, it tells us about the anticipated negative impact at the community level." 
  
##### CCVI Data
* The data is organized by state and county (as well as census tract data), and I decided to summarize the state data at the regional level. I used the regions West, Southwest, Midwest, Southeast, and Northeast to group the states as defined by numerous sources including [National Geographic][2].

* The overall CCVI is a composite of six themes / vulnerability indicators. Below you can see an interactive plot that allows you to view the vulnerability scores for each indicator across U.S. regions, with a higher score meaning higher vulnerability in case there were states with parituclalry high or low vulnerability that may stand out from the region and influence the average. 


```{r eruptions, echo=FALSE}
inputPanel(
 selectInput(inputId = "variable", label = "Indicator:", 
                    choices = 
                      c("Overall CCVI Score" = "ccvi_score",
                        "Socioeconomic Status" = "theme_1",
                        "Household Composition & Disability" = "theme_2",
                        "Minority Status & Language" = "theme_3",
                        "Housing Type & Transportation" = "theme_4",
                        "Epidemiologic Factors" = "theme_5",
                        "Healthcare System Factors" = "theme_6"), multiple = FALSE),
        
        helpText("View the overall scores, or choose from the six themes / indicators that were used to calculate the Overall CCVI Score.", 
                 "A higher score indicates higher vulnerability.")
)

renderPlot({
  # plot ccvi mean / median side-by-side
      ggplot(data = rg_data, aes(x = region, y = unlist(rg_data[,input$variable]), fill = value)) + 
        geom_col(position = position_dodge(), color = 'black', lwd = 0.8) + 
        scale_fill_brewer(palette="Greens") +
        geom_text(aes(label=unlist(rg_data[,input$variable])), position = position_dodge(width = 0.9),
                  vjust=-0.5, size=4) +
        theme_minimal() +
        ylab('Scores') +
        xlab('U.S. Region') +
        ggtitle('Mean / Median Scores across U.S. Regions')
})
```


##### Results
* The results show evidence that communities in the Southeast and Southwest have the highest overall vulnerability, and suggests that those are the regions where more support / resources should be allocated. 

##### Conclusion
* Based on the indicator descriptions given on the Surgo Foundation's [website][1], and evaluating the different indicator scores, it appears these regions were vulnerable due to a variety of factors such as lower income, higher unemployment, larger elderly population, racial marginalization, and other underlying health conditions. It is also notable that the Northeast has the lowest overall score despite New York being called the [epicenter of the pandemic][3] at one point.

[1]: https://precisionforcovid.org/ccvi "CCVI"
[2]: https://www.nationalgeographic.org/maps/united-states-regions/#:~:text=A%20common%20way%20of%20referring,West%2C%20Southeast%2C%20and%20Midwest. "National Geographic"
[3]: https://www.wbur.org/onpoint/2020/04/13/new-york-city-epicenter-pandemic "Epicenter"
[4]: https://github.com/simonminasse/ccvi_regions "Github"

##### Sources
1. CCVI. https://precisionforcovid.org/ccvi
2. National Geographic. https://www.nationalgeographic.org/maps/united-states-regions/#:~:text=A%20common%20way%20of%20referring,West%2C%20Southeast%2C%20and%20Midwest.
3. NYC Epicenter. https://www.wbur.org/onpoint/2020/04/13/new-york-city-epicenter-pandemic 

Feel free to check out the github repo associated with this project [here][4].