---
title: "CCVI across U.S. Regions"
author: "Simon Minasse"
date: "9/26/2020"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("/Users/simonminasse/GitAndR/ccvi_regions")

library(tidyverse)
library(readxl)
library(RColorBrewer)
library(wesanderson)

load('/Users/simonminasse/GitAndR/ccvi_regions/regional_plots.RData')

library(shiny)
```

* Over the past few months I have come across a lot of interesting information and 
resources relating to the current pandemic. One of the things I have noticed in particular is the amount
of publicly accessible data relating to COVID-19. As a result, I felt it was a good idea to use at least 
one of these datasets to present information in a useful way, and sharpen my data skills at the same time.

* The dataset I chose to work on was the [Surgo Foundation's COVID-19 Community Vulnerability Index (CCVI)][1]. The CCVI is a resource that helps to identify areas in the U.S. where support is most needed. As they put it, 
  "The CCVI is not designed to predict which individuals will become infected with coronavirus – instead, it   tells us about the anticipated negative impact at the community level." 
They have data at the state, and county levels, and I decided to summarize the state data at the regional level. I used the regions West, Southwest, Midwest, Southeast, and Northeast to group the states as defined by numerous sources includiong [National Geographic][2].

* The overall CCVI is a composite of six themes / vulnerability indicators. Below you can see an interactive plot that allows you to view the vulnerability scores for each indicator across U.S. regions. 
* I chose to show both the mean and median scores for states in each region because the commonly-used mean may be skewed by outliers, and if such was the case, this visualization would help make it clear. 


```{r eruptions, echo=FALSE}
inputPanel(
 selectInput(inputId = "variable", label = "Indicator:", 
                    choices = 
                      c("Overall CCVI Score" = "ccvi_score",
                        "Socioeconomic Status" = "theme_1",
                        "Household Composition & Disability" = "theme_2",
                        "Minority Status & Language" = "theme_3",
                        "Housing Type & Transportation" = "theme_4",
                        "Epidemiologic Factors" = "theme_5",
                        "Healthcare System Factors" = "theme_6"), multiple = FALSE),
        
        helpText("View the overall scores, or choose from the six themes / indicators that were used to calculate the Overall CCVI Score.", 
                 "A higher score indicates higher vulnerability.")
)

renderPlot({
  # plot ccvi mean / median side-by-side
      ggplot(data = rg_data, aes(x = region, y = unlist(rg_data[,input$variable]), fill = value)) + 
        geom_col(position = position_dodge(), color = 'black', lwd = 0.8) + 
        scale_fill_brewer(palette="Greens") +
        geom_text(aes(label=unlist(rg_data[,input$variable])), position = position_dodge(width = 0.9),
                  vjust=-0.5, size=4) +
        theme_minimal() +
        ylab('Scores') +
        xlab('U.S. Region') +
        ggtitle('Mean / Median Scores across U.S. Regions')
})
```

* The results show evidence that communities in the Southeast and Southwest have the highest overall vulnerability, and suggests that those are the regions where most support should be allocated. Based on the description given on the Surgo Foundation's [website][1], these regions were vulnerable due to factors such as lower income, higher unemployment, larger elderly population, being racially marginalized, and having other underlying health conditions. It is notable that the Northeast has the lowest score despite New York being called the [epicenter of the pandemic][3] at one point. This suggests that areas like New York are better able to respond to the pandemic compared to other areas in the country.

* With regards to sharpening my skillset, this was a good exercise. Although I have not built an RShiny App up to this point, I do have a lot of experience working in R so the learning curve wasn't too bad. I'm sure being able to build this type of visualization will come in handy for future analyses. If you are curious about how I did this, feel free to check out the github repo associated with this project here (link).


[1]: https://precisionforcovid.org/ccvi "CCVI"
[2]: https://www.nationalgeographic.org/maps/united-states-regions/#:~:text=A%20common%20way%20of%20referring,West%2C%20Southeast%2C%20and%20Midwest. "National Geographic"
[3]: https://www.wbur.org/onpoint/2020/04/13/new-york-city-epicenter-pandemic "Epicenter"

